RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-10-07 doc/bioinformatics2013/RVD2.tex
updates to manuscript

\usepackage{setspace}
%\doublespacing

@@

Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}. Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}. As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}. NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis}.

The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology. We present the hierarchical Bayesian model in Section X and a Bayesian hypothesis test to detect mutations in Section X. In Section X we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions. We compare our algorithm to the most accurate methods available to date in Section X. Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.

@@

\subsection{Model Structure}

@@

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Hypothesis Test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

@@

@@

\subsection{Performance with read depth}

@@

\subsection{Empirical performance compared with other algorithms}

@@

\subsection{Pooled Multiple-Sclerosis Samples}

@@

We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA. All position information is referenced to human genome reference hg19. One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts. This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\cite{HahnCho2011}. The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts. This variant has not yet been shown to be associated with any known disease. The IL2RA variant is at chr10:6061965 A$>$T and has a MAF  of 3\% and 5\% in the two cohorts. This variant has not yet been shown to be associated with any known disease.

@@

\section{Comparison to other algorithms}\label{sec:comparison}
\begin{figure}[h]
\begin{center}
\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
\caption{Comparison of RVD2 with other variant calling algorithms (MCC).}
\label{fig:comparison_mcc}
\end{center}
\end{figure}

>>>>

\usepackage{setspace}
\doublespacing

@@

Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}. Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}. As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}. NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.

@@

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.

The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology. We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}. In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions. We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}. Finally, in Section~\ref{sec:ms} we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.

@@

\subsection{Model Structure}\label{sec:model_structure}

@@

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Hypothesis Test}\label{sec:hypothesis_test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

@@

The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the Matthews correlation coefficient.
\begin{equation}
	\tau^* = \arg\max_\tau \text{MCC}(\tilde{n}, \text{MAF})
\end{equation}

@@

\subsection{Performance with read depth}\label{sec:read_depth}

@@

\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}

@@

\subsection{Pooled Multiple-Sclerosis Samples}\label{sec:ms}

@@

We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA. All position information is referenced to human genome reference hg19. One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts. This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\citep{Hahn:2009bi}. The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts. This variant has not yet been shown to be associated with any known disease. The IL2RA variant is at chr10:6061965 A$>$T and has a MAF  of 3\% and 5\% in the two cohorts. This variant has not yet been shown to be associated with any known disease.

@@

\section{Comparison to other algorithms}\label{sec:app_comparison}
\begin{figure}[h]
\begin{center}
\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
\caption{Comparison of RVD2 with other variant calling algorithms (MCC).}
\label{fig:comparison_mcc}
\end{center}
\end{figure}
